### 2019-03-03 paper/sex_selection_spacing-ver2.tex
Minor edits
Make all code and paper run without breaking

\usepackage{amsmath}
\usepackage{fontspec}
\setromanfont[Ligatures=TeX]{TeX Gyre Pagella}
\usepackage{unicode-math}
\setmathfont{TeX Gyre Pagella Math}
\usepackage[title]{appendix}
\usepackage[margin=1.0in]{geometry}
\usepackage[figuresleft]{rotating}
\usepackage[longnamesfirst]{natbib}
\usepackage{dcolumn}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage[flushleft]{threeparttable}
\usepackage{setspace}
\usepackage[justification=centering]{caption}
\usepackage[font=scriptsize]{subfig}
\usepackage[xetex,colorlinks=true,linkcolor=black,citecolor=black,urlcolor=black]{hyperref}
\usepackage{adjustbox}
\usepackage{xfrac}

@@

\date{November 2018}

@@

Strong son preference is typically associated with shorter birth spacing
in the absence of sons, but access to sex selection has the potential to
reverse this pattern because each abortion extends spacing by six to
twelve months. 
I introduce a statistical method that simultaneously
accounts for how sex selection increases the spacing between 
births and the likelihood of a son. 
Using four rounds of India's National
Family and Health Surveys, I show that, except for first births,
the spacing between births increased substantially over the last four
decades, with the most substantial increases among women most 
likely to use sex selection.
Specifically, well-educated women with no boys now
exhibit significantly longer spacing and more male-biased sex ratios
than similar women with boys. 
Women with no education still follow the standard
pattern of short spacing when they have girls and little evidence of sex
selection, with medium-educated women showing mixed results. 
Finally,
sex ratios are more likely to decline within spells at lower parities,
where there is less pressure to ensure a son, and more likely to
increase or remain consistently high for higher-order spells, where the
pressure to provide a son is high.

@@

Parents' spacing of births has long served as a measure of son preference 
\citep{Leung1988}.
Before prenatal sex determination became available, the only recourse for 
parents who wanted a son---but did not yet have one---was to have the next 
birth sooner.
Son preference is therefore often associated with shorter spacing after the 
births of girls than boys 
\citep{Das1987,Rahman1993,Pong1994,Haughton1996,Arnold1997,Soest2012,Rossi2015}.
Shorter spacing is, in turn, associated with worse health outcomes for girls 
and mothers 
\citep{arnold98,Conde-Agudelo2000,Whitworth2002,Razzaque2005,Rutstein2005,Conde-Agudelo2006}.%
\footnote{
Parents are also more likely to cease childbearing after the birth of 
a son than after a daughter 
\citep{repetto72,ben-porath76b,Das1987,Arnold1997,arnold98,clark00,dreze01,filmer09,Basu2010,Altindag2016}.
}

However, the introduction of prenatal sex determination fundamentally changed the 
relationship between son preference and birth spacing.
Couples, who had shorter birth spacing before because of son preference,
now have access to prenatal sex determination and sex-selective abortions,
and each abortion increases birth spacing by six months to a year.
The increase consists of three parts.
First, starting from the time of the abortion, the uterus needs at 
least two menstrual cycles to recover;  otherwise, the likelihood 
of spontaneous abortion increases substantially \citep{zhou00b}.
The second part is the waiting time to conception, which is between 
one and six months \citep{Wang2003}.
Finally, sex determination tests are reliable only from three months 
of gestation onwards.

As a result, we now have a situation where we may observer \emph{longer} spacing
for families with daughters than for families with sons, precisely because
strong son preference leads them to use sex selection.
Working in the opposite direction, couples with strong son preference
will likely still try to conceive earlier in the absence of sons. 
They may even shorten the time until conception, knowing that they might
have to go through multiple pregnancies and abortions before they
conceive a son.
To further complicate matters, we can still observe shorter birth spacing 
after the births of daughters as a representation of son preference for 
families who---for one reason or another---do not use prenatal sex selection.

Spacing, by itself, can therefore no longer be used as a direct measure of 
son preference, but understanding birth spacing remains a critical undertaking.
First, 
birth spacing is still useful in understanding son preference, 
if combined with the likelihood of observing a boy or a girl.
Second, 
the duration between births may be an important factor in parents' 
decisions, either for preference or economic reasons.
It is, for example, possible that even parents with strong son preference 
may reverse their decision to use prenatal sex determination---and carry 
the next pregnancy to term whether male or female---as the duration from 
the previous birth becomes sufficiently long.

At a broader level, we know less about what determines spacing behavior in 
developing countries than in developed countries.%
\footnote{
Even the effects of access to contraceptive on spacing are uncertain.
On the one hand, access to contraceptives allows women to avoid too 
short spacing between birth, which would increase birth spacing.
On the other hand, increased reliability of access and effectiveness of
contraceptives can lead to shorter spacing between births 
\citep{Keyfitz1971,Heckman1976}.
With less reliable contraception parents choose a higher level 
of contraception, which results in longer spacing, to avoid having 
too many children by accident.
But, as contraception becomes more effective parents can more
easily avoid future births, allowing them to reduce the spacing 
between births without having to worry about overshooting their 
preferred number of children.
This idea may explain why better-educated women have
shorter spacing than less educated women in some instances
 \citep{Tulasidhar1993,Whitworth2002}.
These two counteracting effects may explain why finding statistically 
significant effects of contraception use on birth spacing is 
difficult \citep{Yeakey2009}.
}
With increasing numbers of women entering the labor force in developing
countries, understanding how couples make timing decisions will be necessary
for the design of suitable policies \citep{Portner2018}.
Furthermore, if spacing does affect health outcomes for mother and children, it is 
essential to understand what drives changes in spacing.
We know, for example, that health outcomes for girls appear to improve
in the presence of sex selection \citep{Lin2014,Hu2015}.
It is possible that these improvements are an unintended side-effect of
the longer spacing that arises from sex-selective abortions, rather than
because the smaller number of girls makes them more valued as is often
assumed.

In this paper, I introduce and apply a novel empirical method that directly incorporates 
the effects of sex-selective abortions on 
the duration between births
\emph{and} 
the likelihood of a son. 
The method can be used to analyze both situations with and without prenatal
sex selection.
My proposed method allows for the time since the previous 
birth to affect the decision on sex selection.
By examining under what circumstances sex selection decisions change with 
spacing, we can draw a more nuanced picture of the degree of son preference.

% [Why India?]
I apply the method to birth histories of Hindu women, using data from 
India's National Family and Health Surveys (NFHS), covering the period 
1972 to 2016. 
India is a particularly compelling case.
On the one hand, India has seen dramatic increases in the males-to-females ratio 
at birth over the last three decades as access to prenatal sex determination 
expanded 
\citep{das_gupta97,Sudha1999,Arnold2002,retherford03b,jha06}.%
\footnote{
India is not alone; both China and South Korea saw
significant changes in the sex ratio at birth over the same period 
\citep{Yi1993,park95}.
}
On the other hand, research suggests that son preference in 
India, when measured as ideally having more boys than girls, is decreasing 
over time and with higher education \citep{bhat03,pande07}.
% %
% \footnote{
% This measure of son preference is commonly used in the literature. 
% See, for example, \citet{clark00}, \citet{Jensen2009}, and \cite{Hu2015}.
% }

There have also been substantial changes in access and legality of
prenatal sex determination in India over the period covered.
Abortion has been legal in India since 1971 and still is.
The first reports of sex determination appeared around 
1982--83 \citep{Sudha1999,bhat06,Grover2006}.
The number of clinics quickly increased, and knowledge about sex selection 
became widespread after a senior government official's wife aborted a 
fetus that turned out to be male \citep[p.\ 598]{Sudha1999}.
In 1994, the Central Government passed the Prenatal Diagnostic Techniques 
(PNDT) Act, making determining and communicating the sex of a fetus illegal.%
\footnote{
Details about the act are at \href{http://pndt.gov.in/}{http://pndt.gov.in/}.
The number of convictions has been low.
It took until January 2008 for the first state, Haryana, to reach five convictions.
Hence, private clinics apparently operate with little risk of legal action 
\citep{Sudha1999}.
Furthermore, there is little evidence that bans like this significantly
affect sex ratios \citep{Das-Gupta2016}.
}
% Hence, the data make it possible to show how the spacing between births and
% sex ratios have changed with the introduction of prenatal sex determination,
% and whether the ban affected the relationship between birth spacing and sex 
% ratios.

% [Results]
There are four main results.
First, there has been a general increase in the length of spacing between births
over the four decades covered by the data.
The exception is for first births, where the median duration has either
remained the same or slightly declined, although this hides a
significant compression of the variation in spell length.
Second, the most substantial increase in spacing is for the women who
are most likely to use sex selection.
Specifically, among the best-educated women, those with no boys now has 
significantly longer spacing---and a more male-biased sex ratio---than 
similar women with boys.
Third, women with no education still follow the standard pattern of
short spacing when they have girls and little evidence of the use of sex
selection.
In other words, these women adhere to a strategy where they achieve a
son through higher fertility rather than the use of sex selection.
Finally, sex ratios are the most likely to decline within spells at
lower-order spells, where the pressure to provide a son is smaller, and
are more likely to increase or remain consistently high for higher-order
spells, where the pressure to ensure a son is high.

@@

The unit of analysis is a spell, the period from one birth---or for the first spell from 
marriage---to the next birth.
The model is a discrete time, non-proportional, competing risk hazard model with two exit 
states: either a boy or a girl is born.
The standard approach in the spacing literature is to use a single-exit model, but with
access to sex selection the sex of the next child is not random and parents' choices 
will impact the spacing to a girl or a boy differentially.%
\footnote{
\cite{Merli2000} used a discrete hazard model to examine whether 
there were under-reporting of births in rural China, although they 
estimated separate waiting time regressions for boys and girls.
}
Both the non-randomness and differential spacing can be captured with a competing risk set 
up.

I use a non-proportional hazard specification, which allows the shape of the hazard 
functions to vary across groups, rather than the standard proportional hazard function 
for three reasons.
First, it is unlikely---even in the absence of prenatal sex determination---that the 
effect of household characteristics such as the sex composition of previous births have 
the same effect throughout the entire spell.%
\footnote{
Assuming that the effect of sex composition is the same throughout a spell is especially 
problematic for higher-order spells where different sex composition of previous births can 
lead to substantial differences both in the likelihood of progressing to the next birth and
how soon couples want their next child if they are going to have one.
}
Second, the introduction of prenatal sex determination likely exacerbates any bias 
from the proportionality assumption since sex-selective abortion use varies across groups,
which affects birth spacing, and because a household's use of sex selection may vary 
within a spell, which means that the effects of covariates vary within the spell as well.
Finally, the use of a non-proportional specification mitigates any potential effects 
of unobserved heterogeneity when used in conjunction with a flexible baseline hazard 
\citep{Dolton1995}.

@@

\input{../tables/bootstrap_duration_sex_ratio_low.tex}

@@

\subsection{Middle Education Women}

\input{../tables/bootstrap_duration_sex_ratio_med.tex}

@@

\input{../tables/bootstrap_duration_sex_ratio_high.tex}

@@

\begin{figure}[htpb]
\centering
\setcounter{subfigure}{-1}
\subfloat[Rural Women with No Education]{
    \begin{minipage}{0.48\textwidth}
        \captionsetup[subfigure]{labelformat=empty,position=top,captionskip=-1pt,farskip=-0.5pt}
        \subfloat[Probability of no birth yet]{\includegraphics[width=\textwidth]{spell1_low_urban_pps}} 
        \captionsetup[subfigure]{labelformat=parens}
    \end{minipage}
} 
\setcounter{subfigure}{0}
\subfloat[Urban Women with 8 or More Years of Education]{
    \begin{minipage}{0.48\textwidth}
        \captionsetup[subfigure]{labelformat=empty,position=top,captionskip=-1pt,farskip=-0.5pt}
        \subfloat[Probability of no birth yet]{\includegraphics[width=\textwidth]{spell1_high_urban_pps}} 
        \captionsetup[subfigure]{labelformat=parens}
    \end{minipage}
} 
\caption{
Survival curves conditional on progression to first birth; start point is month of marriage
}
\label{fig:pps_spell1}
\end{figure}

@@

\subsection{Does Sex Selection Change within Spells?}

The above results show that length of birth spacing increased 
more for women with only girls than for other sex compositions
as sex ratios became more male-biased, presumably because
of increased use of sex selection.
Neither the median spacing or the conditional survival curves,
however, can tell us whether sex selection varies \emph{within}
a spell or provide any information about the likelihood of
exiting a spell with a birth.
To understand both, I now turn to the standard survival curves
and predicted sex ratios by spell duration.

\begin{figure}[htpb!]
\centering
\rotatebox[origin=c]{90}{\footnotesize{Urban}}
\setcounter{subfigure}{-2}
\subfloat[1995--2004]{
   \begin{minipage}{0.39\textwidth}
       \captionsetup[subfigure]{labelformat=empty,position=top,captionskip=-1pt,farskip=-0.5pt}
       \subfloat[Prob.\ boy (\%)]{\includegraphics[width=\textwidth]{spell2_g3_high_urban_g_pc}}\\
       \subfloat[Prob.\ no birth yet]{\includegraphics[width=\textwidth]{spell2_g3_high_urban_g_s}}
       \captionsetup[subfigure]{labelformat=parens}
   \end{minipage}
}
\setcounter{subfigure}{-1}
\subfloat[2005--2016]{
   \begin{minipage}{0.39\textwidth}
       \captionsetup[subfigure]{labelformat=empty,position=top,captionskip=-1pt,farskip=-0.5pt}
       \subfloat[Prob.\ boy (\%)]{\includegraphics[width=\textwidth]{spell2_g4_high_urban_g_pc}}\\
       \subfloat[Prob.\ no birth yet]{\includegraphics[width=\textwidth]{spell2_g4_high_urban_g_s}}
       \captionsetup[subfigure]{labelformat=parens}
   \end{minipage}
} 
\\
\rotatebox[origin=c]{90}{\footnotesize{Rural}}
\setcounter{subfigure}{0}
\subfloat[1995--2004]{
   \begin{minipage}{0.39\textwidth}
       \captionsetup[subfigure]{labelformat=empty,position=top,captionskip=-1pt,farskip=-0.5pt}
       \subfloat[Prob.\ boy (\%)]{\includegraphics[width=\textwidth]{spell2_g3_high_rural_g_pc}}\\
       \subfloat[Prob.\ no birth yet]{\includegraphics[width=\textwidth]{spell2_g3_high_rural_g_s}}
       \captionsetup[subfigure]{labelformat=parens}
   \end{minipage}
}
\setcounter{subfigure}{1}
\subfloat[2005--2016]{
   \begin{minipage}{0.39\textwidth}
       \captionsetup[subfigure]{labelformat=empty,position=top,captionskip=-1pt,farskip=-0.5pt}
       \subfloat[Prob.\ boy (\%)]{\includegraphics[width=\textwidth]{spell2_g4_high_rural_g_pc}}\\
       \subfloat[Prob.\ no birth yet]{\includegraphics[width=\textwidth]{spell2_g4_high_rural_g_s}}
       \captionsetup[subfigure]{labelformat=parens}
   \end{minipage}
}
\caption{Predicted probability of having a boy and probability of 
no birth yet from nine months after first birth for women with a girl
as their first child and eight or more years of education 
for the 1995--2004 and 2005--2016 periods
} 
\label{fig:decreasing_sex_ratio} 
\end{figure}

>>>>

\usepackage{amsmath}
\usepackage{fontspec}
\setromanfont[Ligatures=TeX]{TeX Gyre Pagella}
\usepackage{unicode-math}
\setmathfont{TeX Gyre Pagella Math}
\usepackage[title]{appendix}
\usepackage[margin=1.0in]{geometry}
\usepackage[figuresleft]{rotating}
\usepackage[longnamesfirst]{natbib}
\usepackage{dcolumn}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage[flushleft]{threeparttable}
\usepackage{setspace}
\usepackage[justification=centering]{caption}
\usepackage[font=scriptsize]{subfig}
\usepackage[xetex,colorlinks=true,linkcolor=black,citecolor=black,urlcolor=black]{hyperref}
\usepackage{adjustbox}
\usepackage{xfrac}
\usepackage{placeins}

@@

\date{March 2018}

@@

[TK rewrite with region results]

Strong son preference is typically associated with shorter birth spacing
in the absence of sons, but access to sex selection has the potential to
reverse this pattern because each abortion extends spacing by six to
twelve months. 
I introduce a statistical method that simultaneously accounts for how sex selection 
increases both the spacing between births and the likelihood of a son. 
Using the four rounds of India's National Family and Health Surveys, I show that
the spacing between births increased substantially over the last four
decades, with the most substantial increases among women most 
likely to use sex selection.
Specifically, well-educated women with no boys now
exhibit significantly longer spacing and more male-biased sex ratios
than similar women with boys. 
Women with no education still follow the standard
pattern of short spacing when they have girls and little evidence of sex
selection, with medium-educated women showing mixed results.

@@

Birth spacing has long served as a measure of son preference, with strong
son preference typically associated with shorter birth intervals the
fewer sons a family has \citep{ben-porath76b,Leung1988}.
However, the introduction of prenatal sex determination has the potential to
fundamentally change the relationship between son preference and birth spacing.
The stronger the son preference, the more likely couples are to resort to sex-selective 
abortions, and each abortion automatically increases the interval between births by six 
months to a year.%
\footnote{
The increase consists of three parts.
First, starting from the time of the abortion, the uterus needs at 
least two menstrual cycles to recover;  otherwise, the likelihood 
of spontaneous abortion increases substantially \citep{zhou00b}.
The second part is the waiting time to conception, which is between 
one and six months \citep{Wang2003}.
Finally, sex determination tests are reliable only from three months 
of gestation onwards.
}
As a result, we may now observe \emph{longer} average spacing for families with the
strongest son preference, precisely because it makes them more likely to use sex selection.
% To further complicate matters, families with son preference who---for one reason or 
% another---do not use prenatal sex selection may continue to have shorter birth spacing 
% than those with a lower preference for sons.

In this paper, I examine how birth spacing in India has changed over time and across 
groups with the introduction of sex selection.
I introduce and apply an empirical method that directly incorporates the effects of 
sex-selective abortions on the duration between births 
\emph{and} 
the likelihood of a son. 
The method can be used to analyze both situations with and without prenatal
sex selection.
I apply the method to birth histories of Hindu women, using data from the four
India's National Family and Health Surveys (NFHS), covering the period 
1972 to 2016.

% [Why India?]
India is a particularly compelling case.
India has a long history of male preference, especially in the northern states 
\citep{Kishor1993,murthi95,arnold98}.
A significant proportion of Indian families exhibit a son preference based 
differential stopping behavior, where, for a given number of children, they are more 
likely to have an additional child if they have not yet reached their preferred number of 
sons \citep{repetto72,Das1987,Arnold1997,arnold98,clark00,Basu2010,Barcellos2014}.%
\footnote{
Other countries show a similar pattern
\citep[see, for example,][]{larsen98,filmer09,Altindag2016}.
}
Moreover, mortality risk is higher for females than males for most ages groups, and
there has been an almost continuous increase in India's overall ratio of males to females 
over the last century \citep{dyson01,Navaneetham2011,Bongaarts2015}.

% [What do we know about use of sex selection?]
Over the last three decades, India has seen dramatic increases in the males-to-females 
ratio at birth as access to prenatal sex determination expanded 
\citep{das_gupta97,Sudha1999,Arnold2002,retherford03b,jha06,Guilmoto2012,Portner2015b}.
Simultaneously, fertility has declined substantially, to the point where it is now at, or 
even below, replacement in some areas 
\citep{Guilmoto2013,Dharmalingam2014,International-Institute-for-Population-Sciences-IIPS2017}.
This decline in fertility, combined with a continued strong preference for sons, has 
intensified the pressure to use sex selection 
\citep{das_gupta97,Guilmoto2009a,Bongaarts2013,Jayachandran2017}.%
\footnote{
In some instances, however, the fertility decline appears to have created a stronger 
aversion to daughters rather than a stronger preference for son \citet{DiamondSmith2008}.
}
At household level the strongest predictor of the use of sex selection is the sex 
composition of previous children \citep{retherford03b,jha06,abrevaya09,Portner2015b}.
For families without a son, the higher the parity, the higher is the probability
of having a son as the next birth.
The use of sex selection also increases with socioeconomic status, especially 
education, and is more prevalent in cities than in rural areas, both possibly because 
of lower desired fertility \citep{retherford03b,jha06,Portner2015b}.

% [Why spacing?] 
There are four main reasons for examining birth spacing and how it changes with 
access to sex selection.
First, researchers have made extensive use of birth spacing as a measure of son preference, 
and it is critical to understand to what extent spacing is still useful as a measure of son 
preference and develop new methods if needed.
In India, the sex of the last-born child significantly impacts the duration to next birth,
with the expected birth interval about three percent longer if the last-born was a son than
if it was a daughter \citep{Bhalotra2008,Kumar2016}.
Similarly, ethnic Indians in South Africa also show a longer duration after the birth of a 
son than after a daughter \citep{Gangadharan2003}.%
\footnote{
Ethnic Indians in Malaysia do, however, show little evidence of son preference, although
the sample sizes are small \citep{Pong1994}.
}
In both Bangladesh and Vietnam the more boys a family has the longer the expected duration 
to the next birth \citep{Haughton1995,Haughton1996,Rahman1993,Soest2018}.
Outside Asia, the evidence is more mixed.
North Africa shows shorter spacing in the absence of sons, while a similar pattern does
not exist in Sub-Saharan Africa \citep{Rossi2015}.%
\footnote{
A possible exception is Senegal, where women with a higher risk of widowhood show
shorter birth spacing until they have secured a son \citep{Lambert2016}.
}

Second, birth spacing can have a substantial impact on children's outcomes.
Most of the research finds a negative impact of short spacing on child health, although 
identifying the causal mechanisms have proven more difficult
\citep{Conde-Agudelo2006,Conde-Agudelo2012}.
There are three pathways that birth intervals can affect health.
First, there is maternal depletion, where the mother does not have sufficient time to 
recover after her last pregnancy before she become pregnant again.%
\footnote{
This depletion can take the form of nutrition, folate, or cervical insufficiency.
}
Second, disease transmission both from mother to child and between children close in age.
Third, sibling competition, where siblings close in age compete for scarce resources and
parental care.
Which mechanism dominates is of particular interest when son preference drives spacing 
decisions.
Both the maternal depletion and the transmission of disease explanations predominately
affect the subsequent child and the mother.
Only sibling competition has the potential to negatively affect the older child.

In India, there is evidence that both the prior and the subsequent child have worse health 
outcomes and a higher likelihood of dying the shorter the interval between them, although 
the effect is not symmetrical
\citep{Whitworth2002,Bhargava2003,Rutstein2005,Maitra2008,Makepeace2008,Ghosh2018}.%
\footnote{
\cite{Bhalotra2008}, however, find that birth spacing only explains a limited amount
of neonatal mortality.
}
The increased mortality risk for prior children may come from shorter breastfeeding of 
girls to ensure that the mother can conceive again quicker, and with poor water
quality this shorter breastfeeding may result both in higher risk of dying and shorter
stature if the child survives \citep{Jayachandran2011,Jayachandran2017a}.
In other countries, focus has been mostly on the subsequent birth, but there is evidence 
in countries as diverse as Bangladesh, Brazil, El Salvadore, and Pakistan that
short intervals between births have adverse health effects
\citep{Cleland1984,Curtis1993,Davanzo2008,Gribble2009,Saha2013}.

The potential effects of spacing are not restricted to early outcomes, although
there is less evidence on longer-term effects, and most comes from Western countries.
In the US, for example, a longer spacing between births increases the older sibling's
test scores on the Peabody Individual Achievement Test, although there is no
statistically significant effect on the younger sibling \citep{Buckles2012}.
In line with this, close spacing appears to increase the risk of dropping out of
high school and lower the probability of attending a post-secondary school in both the
US and Sweden \citep{Powell1993,Pettersson-Lidbom2009}.
These results are, however, not uniformly supported, with recent results from Sweden 
showing no relationship between birth spacing, even if very short, and outcomes such as 
years of education completed, earnings, and unemployment \citep{Barclay2017}.

Third, the mother may also be affected by the length of the birth intervals.
The effect of spacing on mothers have received less attention than the effect on children, 
and reviews of the literature show mixed results on mother's anthropometric status 
\citep{Dewey2007,Conde-Agudelo2012}.
For India, there is, however, evidence that women with first-born girls are both
more likely to have short birth spacing and more likely to have anemia, possibly as
the result of the short spacing resulting in maternal depletion \citep{Milazzo2018}.
In Western countries, having a longer interval between births appear to have positive
effects on long-term labor market outcomes, such as participation and 
income \citep{Gough2017,Karimi2014}.

Finally, we know less about what determines spacing behavior in developing countries than 
in developed countries, and with declining fertility and increasing numbers of women 
entering the labor force in developing countries, understanding how couples make timing 
decisions will be necessary for the design of suitable policies \citep{Portner2018}.%
\footnote{
It is clear that birth spacing does respond to policies in both developed and 
developing countries \citep{Pettersson-Lidbom2009,Todd2012,Meckel2015,Ghosh2018}.
}
For example, the evidence on the effects of education and socio-economic status on
birth intervals is mixed, with some research showing either a positive association, no
association, or one that is changing over time 
\citep{Tulasidhar1993,Whitworth2002,Bhalotra2008,Kim2010,Soest2018}.

Even the effects of access to contraceptive on spacing are unclear.
On the one hand, access to contraceptives allows women to avoid too short spacing between 
birth, which would increase birth spacing.
On the other hand, increased reliability of access and effectiveness of contraceptives can 
lead to shorter intervals between births  \citep{Keyfitz1971,Heckman1976}.
With less reliable contraception parents choose a higher level of contraception, which 
results in longer spacing, to avoid having too many children by accident.
However, as contraception becomes more effective parents can more easily avoid future 
births, allowing them to reduce the spacing between births without having to worry about 
overshooting their preferred number of children.
This idea may explain why better-educated women have shorter spacing than less educated 
women in some instances but not in others.
These two counteracting effects may explain why finding statistically significant effects 
of contraception use on birth spacing is difficult \citep{Yeakey2009}.%
\footnote{
There is evidence in Matlab that the effect of son preference on birth spacing is
stronger in areas with better access to family planning \citep{Rahman1993}.
Other evidence, however, points to families ability to time birth, even in the absence
of modern contraceptives \citep{Jayachandran2011,Alam2018}.
}

TK spacing and decline in fertility (with and without sex selection)

Despite the increasing use of sex selection, the extensive use of spacing to measure son 
preference, and the important impacts of spacing on outcomes for both mother and children  
there has so far been no examination of what effect access to sex selection has had on 
spacing.

TK What questions I am asking

TK How I am answering them

Direct information on the use of sex selection is not available, but it is possible
to compare different periods as a proxy since there has been substantial changes in access 
and legality of prenatal sex determination in India.
Abortion has been legal in India since 1971 and still is.
The first reports of sex determination appeared around 
1982--83 \citep{Sudha1999,bhat06,Grover2006}.
The number of clinics quickly increased, and knowledge about sex selection 
became widespread after a senior government official's wife aborted a 
fetus that turned out to be male \citep[p.\ 598]{Sudha1999}.
In 1994, the Central Government passed the Prenatal Diagnostic Techniques 
(PNDT) Act, making determining and communicating the sex of a fetus illegal.%
\footnote{
Details about the act are at \href{http://pndt.gov.in/}{http://pndt.gov.in/}.
The number of convictions has been low.
It took until January 2008 for the first state, Haryana, to reach five convictions.
Hence, private clinics apparently operate with little risk of legal action 
\citep{Sudha1999}.
Furthermore, there is little evidence that bans like this significantly
affect sex ratios \citep{Das-Gupta2016}.
}
[TK something on split 1995 to 2016 into two groups?]

% [Results - need updating]
There are three main results.
First, there has been a general increase in the length of spacing between births
over the four decades covered by the data.
Second, the most substantial increase in spacing is for the women who
are most likely to use sex selection.
Specifically, among the best-educated women, those with no boys now has 
significantly longer spacing---and a more male-biased sex ratio---than 
similar women with boys.
Third, women with no education still follow the standard pattern of
short spacing when they have girls and little evidence of the use of sex
selection.
In other words, these women adhere to a strategy where they achieve a
son through higher fertility rather than the use of sex selection.

@@

The unit of analysis is a spell, the period from one birth to the next.
The model is a discrete time, non-proportional, competing risk hazard model with two exit 
states: either a boy or a girl is born.
The standard approach in the spacing literature is to use a single-exit model, but with
access to sex selection the sex of the next child is not random and parents' choices 
will impact the spacing to a girl or a boy differentially.%
\footnote{
\cite{Merli2000} used a discrete hazard model to examine whether 
there were under-reporting of births in rural China, although they 
estimated separate waiting time regressions for boys and girls.
}
Both the non-randomness and differential spacing can be captured with a competing risk 
set up.

I use a non-proportional hazard specification, which allows the shape of the hazard 
functions to vary across groups, rather than the standard proportional hazard function 
for three reasons.
First, it is unlikely---even in the absence of prenatal sex determination---that the 
effect of household characteristics such as the sex composition of previous births have 
the same effect throughout the entire spell.%
\footnote{
Assuming that the effect of sex composition is the same throughout a spell is especially 
problematic for higher-order spells where different sex composition of previous births can 
lead to substantial differences both in the likelihood of progressing to the next birth and
how soon couples want their next child if they are going to have one.
}
Second, the introduction of prenatal sex determination exacerbates any bias from the 
proportionality assumption since sex-selective abortion use varies across groups,
which affects birth spacing, and because a household's use of sex selection may vary 
within a spell, which means that the effects of covariates vary within the spell as well.
Finally, the use of a non-proportional specification mitigates any potential effects 
of unobserved heterogeneity when used in conjunction with a flexible baseline hazard 
\citep{Dolton1995}.

@@

\begin{figure}[htpb]
\centering
\caption*{Urban}
\subfloat[all]{\includegraphics[width=0.27\textwidth]{bs_spell2_low_urban_all}} 
\subfloat[all]{\includegraphics[width=0.27\textwidth]{bs_spell3_low_urban_all}} 
\subfloat[all]{\includegraphics[width=0.27\textwidth]{bs_spell4_low_urban_all}} 
\\
\caption*{Rural}
\subfloat[all]{\includegraphics[width=0.27\textwidth]{bs_spell2_low_rural_all}} 
\subfloat[all]{\includegraphics[width=0.27\textwidth]{bs_spell3_low_rural_all}} 
\subfloat[all]{\includegraphics[width=0.27\textwidth]{bs_spell4_low_rural_all}} 
\caption{low education}
\end{figure}

\input{../tables/bootstrap_duration_sex_ratio_low_all.tex}

@@

\subsection{Low to Medium Education Women}

\begin{figure}[htpb]
\centering
\caption*{Urban}
\subfloat[all]{\includegraphics[width=0.27\textwidth]{bs_spell2_med_urban_all}} 
\subfloat[all]{\includegraphics[width=0.27\textwidth]{bs_spell3_med_urban_all}} 
\subfloat[all]{\includegraphics[width=0.27\textwidth]{bs_spell4_med_urban_all}} 
\\
\caption*{Rural}
\subfloat[all]{\includegraphics[width=0.27\textwidth]{bs_spell2_med_rural_all}} 
\subfloat[all]{\includegraphics[width=0.27\textwidth]{bs_spell3_med_rural_all}} 
\subfloat[all]{\includegraphics[width=0.27\textwidth]{bs_spell4_med_rural_all}} 
\caption{med education}
\end{figure}

\input{../tables/bootstrap_duration_sex_ratio_med_all.tex}

@@

\begin{figure}[htpb]
\centering
\caption*{Urban}
\subfloat[all]{\includegraphics[width=0.27\textwidth]{bs_spell2_high_urban_all}} 
\subfloat[all]{\includegraphics[width=0.27\textwidth]{bs_spell3_high_urban_all}} 
\subfloat[all]{\includegraphics[width=0.27\textwidth]{bs_spell4_high_urban_all}} 
\\
\caption*{Rural}
\subfloat[all]{\includegraphics[width=0.27\textwidth]{bs_spell2_high_rural_all}} 
\subfloat[all]{\includegraphics[width=0.27\textwidth]{bs_spell3_high_rural_all}} 
\subfloat[all]{\includegraphics[width=0.27\textwidth]{bs_spell4_high_rural_all}} 
\caption{high education}
\end{figure}

\input{../tables/bootstrap_duration_sex_ratio_high_all.tex}

@@

% \begin{figure}[htpb]
% \centering
% \setcounter{subfigure}{-1}
% \subfloat[Rural Women with No Education]{
%     \begin{minipage}{0.48\textwidth}
%         \captionsetup[subfigure]{labelformat=empty,position=top,captionskip=-1pt,farskip=-0.5pt}
%         \subfloat[Probability of no birth yet]{\includegraphics[width=\textwidth]{spell1_low_urban_pps}} 
%         \captionsetup[subfigure]{labelformat=parens}
%     \end{minipage}
% } 
% \setcounter{subfigure}{0}
% \subfloat[Urban Women with 8 or More Years of Education]{
%     \begin{minipage}{0.48\textwidth}
%         \captionsetup[subfigure]{labelformat=empty,position=top,captionskip=-1pt,farskip=-0.5pt}
%         \subfloat[Probability of no birth yet]{\includegraphics[width=\textwidth]{spell1_high_urban_pps}} 
%         \captionsetup[subfigure]{labelformat=parens}
%     \end{minipage}
% } 
% \caption{
% Survival curves conditional on progression to first birth; start point is month of marriage
% }
% \label{fig:pps_spell1}
% \end{figure}

@@

\FloatBarrier
\section{Conclusion\label{sec:conclusion}}
